Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-08-12 Response_to_Reviewers.tex

\textbf{Response to reviewers}

%Resubmit 15 Aug. 2019\\

\bigskip

We want to thank the two reviewers for their thorough evaluation and discussion of our paper, as well as the challenging questions they asked. We have revised the manuscript in a way that we believe addresses most of the comments. In this response, we first want to address the two main questions that arose for both reviews. The first one (A) relates to the original mis-specification/ambiguities between the social and spatial components of the modelling process. The second one (B) relates to the contribution of the approach in terms of potential use for model makers, for domain experts and in operational contexts.

@@

 \medskip
A. In terms of the ambiguity between the social and spatial component 
\begin{itemize}
\item A.1. We have precised and clarified the argument of the paper (mainly in introduction) by qualifying our contribution, which is to provide an operational framework for studying the influence of geometric structures of space on the results of a geosimulation model. We have thus distinguished two effects of geographical space in simulation modelling: the ones that are part of the model, reflecting for example the role played by distance on decreasing agents' perception or potential social interactions; and the ones which act as geometric structures "on which" the social elements simulated interact, i.e. geometric constraints of the simulation. The first type of effects is addressed by traditional modelling. Our original contribution is to tackle the second type. %\comment[CC]{done in 1.5 and 1.8}

\item A.2. We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e. groups of social agents whose behaviour is constrained by their position in geographical space)%. This term refers to empirical systems, which are the targets of the modelling process)% Mise en commentaire par FL 190812 : je trouve que cela alourdi inutilement la phrase
, \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we use for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model. When we talk about the generator we actually used in the paper, we call it a "density grid generator" because it generates a square grid with heterogeneous density, whereas in all generality a "spatial generator" could potentially generate graph or point structures. %\comment[CC]{done in 1.5} %Changement par FL 190812 Because this is the only geometric structure it generates

@@

 \item A.3. We have modified figure 1 to express the clarified approach visually. Naming the generator parameter "gamma" has helped reduced the confusion that the term 'meta-parameter' was inducing. We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper. %\comment[CC]{done}
\end{itemize}

 \medskip
B. To highlight limitations as well as promising perspectives of improvement, we answer the concern about the genericity of the method and expose its links with the conceptual and empirical domains:

@@

 \begin{itemize}

  \item B.1. We have added a discussion of\comment[FL]{on?} how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities). This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research. % \comment[MLT]{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.3}

  \item B.2. To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in discussion, explaining how our approach allows us to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences). \comment{il faut reprendre ces elements dans l'article}

  \item B.3. As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper)\comment[FL]{pas clair : pour calculer les indicateurs? pour les interactions? à préciser comment sont utilisés ces ego neighborhoods}. We could also\comment[FL]{plutôt : In further researches, we plan to extend the analysis...} in another paper vary the size of grid and its cell shape (cf. MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example points or networks). % \comment{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.1}
\end{itemize}

@@

\medskip

%The paper describes an approach that takes spatial sensitivity analysis explicitly and thoroughly into account. Based on the criticism that most agent-based-modelling and simulation approaches do neglect a careful recognition of the spatial impact on social phenomena, it introduces a “spatial generator” whose task is to generate different possible spatial configurations that are then used as a comparative measurement in socio-spatial simulations to prove the influence of spatial structures. To evaluate the strengths and weaknesses of the use of the spatial generator tool the author(s) applies two classical approaches – Schelling’s segregation model and the Sugarscape model – as benchmark models.

\begin{enumerate}
 % The methodology described is very convincing as it allows the utilisation of a tool that takes the spatial conditions – as causes or effects or intermediates – actually explicitly into account. With this approach, the paper convincingly outlines a pragmatic application by creating typical spatial compositions in urban contexts that is monocentric, polycentric or discontinuous cities.
 	\item \textit{From this methodological perspective of reviewing, it would be helpful to add some more information to approaches/techniques used within the tool of the spatial generator. For example, paragraph 2.8 mentions four dimensions of spatial structure but leave the reader alone with just the terms. Either describe these or omit it.}

   $\rightarrow$ To describe in more details the density grid generator we added a few sentences; and the generator was furthermore better situated among the many possible others in the discussion.
   %ways of making a spatial generator as Young2002 (FLN/MLT).
 	% item \comment[JR]{will add some description on urban form indicators etc.}

@@

 \medskip

  %\item reprendre tout le papier pour éliminer les termes indéfinis "space", "pattern" etc. + en conclu mentionner que ça s'applique aux automates cellulaires (sans agents) et aux modèles écologiques (agents-arbres).
  %\item in conclusion, we need to clarify the answer to the question "do geometric structure influence sociospatial systems dynamics" by saying it depends on the model. ≠ black\&white and we can demonstrate it. The nature of social systems is a given (in the model mu, which is not the contribution of this paper) and we explore the consequence of geometric structures on the spatial expression of social systems.

  \item \textit{Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.} \ldots
  % Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account. These interactions can then be associated with the material and relational dimension of geographical spaces. The question remains: what is the role of geographical systems? In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.

  $\rightarrow$ see comment A. cite Banos, 2012 \comment[FL]{quel était le point exactement avec Banos?}

@@

  $\rightarrow$ We agree with this comment and recall that urban growth processes are not the main focus of the paper, they are here an instantiation of a given $\gamma$ generator, which could have been replaced by others such as network models, procedural modeling, etc. The position and discussion of the paper was amended to precise that (see comment A).

  %it's not the point of the paper, only the metaphore of the spatial generator. + supprimer la ref à "urban growth" dans le texte et dire d'inspiration urban science mais dont seules les grilles finales nous intéressent. On aurait pu utiliser des generator issus de graphes, de percolation etc.
  %The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
  %\item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.

  \medskip

  \item \textit{The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless. The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive. Either it is a trivial fact, simply because polycentric cities have more centres and thus, by nature, more opportunities from which segregation can originate. Or, the results can be used as a starting point to dig deeper in geographical reasons, taking social, economic, cultural, and political, but also topographical issues into consideration. This geographical reasoning, however, is outside of the scope of this paper. }\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
 \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output. Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones? And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper. But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FK]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
  We understand this statements, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer. Instead, we wished to show what kind of results can be obtained using the framework we suggest. If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London. We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results? \comment[FL]{à la relecture de l'article s'assurer que cette prise de recul est bien explicitée}

  \item \comment[FL]{ne pas oublier les trois points ci-dessous!!!}
  \item in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf. comment B) but on the processes, there is nothing obvious about the influence of structures on behavioural spatial patterns (if you imagine a model with no interactions between agents for example) + it's a difference between the model and its potential addition to the real systems. cf. B

@@

  $\rightarrow$ see comment B

  \medskip

  \textit{For instance, it would be useful to have suggestions on how to use this methodology to run empirical models. Suppose to have a nice dataset and a case of a urban context to be modelled. Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?}

  $\rightarrow$ cf. comment B.3
  % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
  it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig. 1). We thank reviewer 2 for highlighting this limitation.

@@

  \medskip

  \item \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}

### 2019-08-12 paper.tex
Update on Overleaf.

\begin{abstract}
Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic geometric structures generator in order to create density grids with which socio-spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change.
\end{abstract}

>>>>

\begin{abstract}
Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic geometric structures generator in order to create density grids with which socio-spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models. We wish to illustrate the potential  add spatial sensitivity analysis during the exploration of their model and that specialist of thethematics would see the opportunity to dig deeper into our results?%Furthermore, this effect is sometimes larger than the effect of parameters' value change.%Comment FL : j'ai supprimé cette dernière phrase qui ne me semblait pas essentielle et rajouté à la place une phrase d'ouverture de Marion.
\end{abstract}
